TRIK Studio: Technical Introduction

### 2017-02-08 2017-trikStudioEnglish/trikStudioEnglish.tex
chapter II is done

\section{Introduction}
Current state of school education in computer science turned out to be quite like Seymour Papert predicted it to be. In 1967 he introduced a virtual Logo turtle, that is used to teach students programming at schools even nowdays. It is less known that Papert in his experiments also used a mechanical robot turtle, that was controlled from a computer ~\cite{papert1980mindstorms}, and it made educational process much more entertaining. Today Papert's ideas are widely spread, a lot of schools are using robots to teach programming (for instance, in Russia robotics is a part of compulsory education program within the Technology course \cite{черёмухин2014внедрение,лучин2016внедрение}). Several robotics educational kits are used, including Lego Mindstorms NXT, Lego Mindstorms EV3 \footnote{LEGO Mindstorms homepage, URL: http://www.lego.com/en-us/mindstorms (accessed: 07.02.2017)}, 
TRIK\footnote{TRIK robotics platform homepage, URL: http://www.trikset.com/ (accessed: 07.02.2017)} etc.

@@

The the first mode the program is being interpreted on a two-dimentional robot model (see~\ref{chapter:2dModel}). Users are able to create 2D-model of the world surrounding the robot out of walls, colored elements and floor markup. According to TRIK Studio users this feature is very useful for initial program debugging, before any interaction with real robots. Our experience shows that this virtual model editor allows to recreate most of fields and obstacle courses used in competitions in robotics. Having such a simulation environment makes possible learning robotics and programming even without having the real robotic kits. There is also an experimental support for V-Rep 3D visual simulation environment~\cite{rohmer2013v}.

Отладка на компьютере с посылкой команд роботу (режим \textit{интерпретации} в терминах среды) удобна для отслеживания поведения программы на целевом устройстве в реальном времени. В режиме интерпретации можно отслеживать значения переменных в соответствующем окне среды (аналогичному, например, окну поддержки отладчика gdb в различных текстовых IDE), а также строить в реальном времени графики значений с датчиков.

Режим генерации кода позволяет перейти от визуального представления программы к текстовому. Тестовый код отображается во встроенном редакторе qscintilla\footnote{https://riverbankcomputing.com/software/qscintilla/intro (дата обращения: 14.05.2016)}, который обладает возможностью полноценного редактора кода (подсветка синтаксиса, автодополнение, подсветка скобок, отмена/повтор и т.д.). В дистрибутив среды входят все необходимые инструменты для построения и передачи программ на робота (набор кросскомпиляторов, WinSCP, Putty и т.д.), поэтому процесс компиляции и взаимодействия с контроллерами роботов остается полностью <<прозрачным>> для пользователей (пользователи-новички до определенного момента даже не догадываются о его существовании).

Пользовательский интерфейс среды представлен на рисунке~\ref{image:TS_interface}. На нем запечатлен момент отладки программы выезда робота из лабиринта на двумерном симуляторе.

\begin{figure*}[ht]
    \centering
    \includegraphics[width=\textwidth]{TS_CF_Labyrinth.png}
    \caption{Интерфейс TRIK Studio}
    \label{image:TS_interface}
\end{figure*}

В режиме двумерной симуляции доступна возможность автоматической проверки задач (см. раздел~\ref{chapter:constraintsChecker}). Программа проверки пишется на внутреннем событийном текстовом языке, файл с сохранением модели симулируемого мира и программы проверки затем может распространяться как задача для учеников. На основе этой системы был запущен курс дистанционного обучения на платформе Stepik\footnote{https://stepik.org/s/7qe3xj4Z  (дата обращения: 14.05.2016)} с видео-лекциями по основам кибернетики, робототехники и использованию среды TRIK Studio, множеством небольших тестов и 20 задачами образовательной робототехники. Каждая задача может быть скачана, решена и проверена в TRIK Studio на локальном компьютере и сдана на сервер, где система проверки запустит решение на своих тестах (подобно формату решения задач на соревнованиях ACM ICPC).

Среда написана на языке C++ с инструментарием Qt\footnote{https://www.qt.io/ru/ (дата обращения: 14.05.2016)}, поэтому является кроссплатформенной (доступны установочные пакеты под Windows, Linux и Mac OS X). При этом, даже несмотря на то, что официальные драйвера для Lego NXT не доступны для Linux и Mac OS X x64, в TRIK Studio существует собственная их реализация, подробнее см. главу~\ref{chapter:communications}. Среда полностью бесплатна, имеет открытый исходный код и распространяется под лицензией Apache License 2.0\footnote{http://www.apache.org/licenses/LICENSE-2.0 (дата обращения: 14.05.2016)}.

Описанные достоинства выгодно выделяют TRIK Studio на фоне всех аналогов. Фактически, среди упомянутых во введении систем, применяемых в школьном образовании, лишь среда 12Blocks обладает приближенной функциональностью, однако генерирует не читаемый код (для Lego NXT), не имеет средств проверки задач, является платной и не русифицирована. Основные недостатки TRIK Studio на данный момент --- это слабая методическая поддержка среды, ограничивающаяся лишь справкой на русском язык, набором примеров и упомянутым курсом дистанционного обучения. Существуют и другие, более мелкие отрицательные особенности, о которых будет упомянуто в соответствующих разделах ниже.

Остаток статьи будет построен следующим образом. В главах~\ref{chapter:controlFlowLanguage} и~\ref{chapter:dataFlowLanguage} будут кратко описаны визуальные языки TRIK Studio и их интерпретаторы. Глава~\ref{chapter:commonArchitecture} дает общее представление об архитектуре среды. Дальнейшие главы будут посвящены конкретным подсистемам TRIK Studio. Общие сведения о реализации механизмов коммуникации с роботами даны в главе~\ref{chapter:communications}. Описания интерпретаторов визуальных языков в системе даны в главе~\ref{chapter:interpreters}. Наиболее интересные детали реализации генераторов в текстовые языки из диаграмм потока управления представлены в главе~\ref{chapter:generators}. В разделе~\ref{chapter:parser} описывается подсистема синтаксического анализа текстовой части языка. В главе~\ref{chapter:2dModel} рассказывается об особенностях реализации подсистемы двумерного имитационного моделирования, далее, в главе~\ref{chapter:constraintsChecker} описан язык описания программ автоматической проверки заданий, исполняемых на этой подсистеме. Наконец, заключительная глава подводит итоги работы.

>>>>

\section{Introduction}
\label{chapter:intro}
Current state of school education in computer science turned out to be quite like Seymour Papert predicted it to be. In 1967 he introduced a virtual Logo turtle, that is used to teach students programming at schools even nowdays. It is less known that Papert in his experiments also used a mechanical robot turtle, that was controlled from a computer ~\cite{papert1980mindstorms}, and it made educational process much more entertaining. Today Papert's ideas are widely spread, a lot of schools are using robots to teach programming (for instance, in Russia robotics is a part of compulsory education program within the Technology course \cite{черёмухин2014внедрение,лучин2016внедрение}). Several robotics educational kits are used, including Lego Mindstorms NXT, Lego Mindstorms EV3 \footnote{LEGO Mindstorms homepage, URL: http://www.lego.com/en-us/mindstorms (accessed: 07.02.2017)}, 
TRIK\footnote{TRIK robotics platform homepage, URL: http://www.trikset.com/ (accessed: 07.02.2017)} etc.

@@

The the first mode the program is being interpreted on a two-dimentional robot model (see Section~\ref{chapter:2dModel}). Users are able to create 2D-model of the world surrounding the robot out of walls, colored elements and floor markup. According to TRIK Studio users this feature is very useful for initial program debugging, before any interaction with real robots. Our experience shows that this virtual model editor allows to recreate most of fields and obstacle courses used in competitions in robotics. Having such a simulation environment makes possible learning robotics and programming even without having the real robotic kits. There is also an experimental support for V-Rep 3D visual simulation environment~\cite{rohmer2013v}.

Debugging using a PC (so called \textit{interpretation mode}) is useful as a next step to see how the program behaves on a robot in real time. In this mode program variables' values can be observed in a special window (similar to how it is done in Watch list windows in almost all textual IDEs) and sensor data can be displayed in a form of graphs.

Code generation mode enables users to move from visual to textual programs. Gerenated code is displayed in an embedded QScintilla-based~\footnote{https://riverbankcomputing.com/software/qscintilla/intro (accessed: 07.02.2017)} text editor, that provides a set of common features of a code editor: syntax highlighting, code audocompletion, undo/redo, brackets highlighting etc. TRIK Studio installation contains everything neede for buiding and uploading programs into a robot (a number of cross-compilers, WinSCP, Putty, etc.), and it allows to make compilation process and communication with robots transparent to TRIK Studio users.

TRIK Studio's user interface is shown on fig.~\ref{image:TS_interface}. It displays process of debugging a program that passes the maze using virtual simulation model.

\begin{figure*}[ht]
    \centering
    \includegraphics[width=\textwidth]{TS_CF_Labyrinth.png}
    \caption{TRIK Studio's user interface}
    \label{image:TS_interface}
\end{figure*}

In 2D simulation mode the automatic task checking feature is available (see Section~\ref{chapter:constraintsChecker}). Task checking program is written in internal event-based textual language, the file containing virtual world model and task checking program can be distributed as a task for students. Based on this feature a remote course has been launched on Stepik platform\footnote{https://stepik.org/s/7qe3xj4Z (accessed: 07.02.2017)} containing video lectures on cybernetics and robotics, numerous small tests and more that 20 tasks on educational robotics. Each task can be downloaded, solved, checked within TRIK Studio on a student's computer, and uploaded to the server where the checking system will run it on its own tests (similar to ACM ICPC coding contests).

TRIK Studio is written in C++ using cross-platform Qt\footnote{https://www.qt.io/ru/ (accessed: 14.05.2016)} framework, there are installers for Windows, Linux and Mac OS X operating systems. Because of the fact that official Lego NXT drivers are not available on Linux and Mac OS X x64, TRIK Studio contains our own implementation for them (see Section~\ref{chapter:communications} for details). The programming environment is completely free to use, open-source and is distributed under Apache License 2.0\footnote{http://www.apache.org/licenses/LICENSE-2.0 (accessed: 07.02.2017)}.

Aforementioned features distinguish TRIK Studio from all other such tools. In fact among all mentioned in Section~\ref{chapter:intro} tools only 12Blocks have a comparable feature set, but it generates complex and unreadable code (for Lego NXT), does not have tools for checking tasks, is not free to use and does not have Russian localization in it. Among the TRIK Studio drawbacks we should mention weak methodological support, limited only by embedded reference guide in Russian, a set of examaples and the remote video course. Less important drawbacks will be mentioned in the following sections.

The rest of the paper is structured as follows. Sections~\ref{chapter:controlFlowLanguage} and~\ref{chapter:dataFlowLanguage} briefly present TRIK Studio visual languages and their interpreters. Section~\ref{chapter:commonArchitecture} overviews the system architecture. Further sections describe separate TRIK Studio subsystems. Section~\ref{chapter:communications} presents robots communication infrastructure. Section~\ref{chapter:interpreters} describes visual language interpreters. The most interesting implementation details of code generators from control-flow language are provided in Section~\ref{chapter:generators}. Section~\ref{chapter:parser} presents textual languages parsing subsystem. Section~\ref{chapter:2dModel} presents implementation details of 2D simulation model sybsystem. Section~\ref{chapter:constraintsChecker} describes task checking language and Section~\ref{chapter:conclusion} concludes the paper.
